---
title: "Let's break some p-values"
author: "Steph and John"
date: "`r Sys.Date()`"
slug: p-values
tags: p-values
categories:
- Bayes
- Frequentist
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```



# Intro banter

__John__: Where do you want to start Steph?

__Steph__: I don't know John. How about we start with something that everyone 
reading this is probably familiar with? Did you want to do some $p$-value bashing?

__John__: Yeah, there are some Bayesians out there who refuse to do hypothesis
testing on principal.

__Steph__: There is a lot of talk about reproducibility, replicability, robustness,
and practices that lead to finding "truth" in the context of scientific practice. 
These issues also com
It might be a good place to start.

__John__: So what do you think

__John__: So a few years ago my wife was rear-ended while driving and the
car was written off so we needed to buy a new car. While we were looking to
get a new car my sister in law asked is (as a joke) "Are you getting a red car?
I heard their fast!"

# Red cars are faster!

Let us imagine a simplified world. 
Suppose that

* There are only two car colours:
<span style="color:red">__red__</span> cars and 
<span style="color:blue">__blue__</span> cars.

* Cars can eiher be <span style="color:purple">__sports__</span> cars and 
<span style="color:green">__normal__</span> cars.

* <span style="color:purple">__Sports__</span> cars
are more likely to be <span style="color:red">__red__</span>
than <span style="color:green">__normal__</span> cars


* <span style="color:purple">__sports__</span> cares
are generally faster than 
<span style="color:green">__normal__</span> cars.

# Directed acyclic graphs

Directed acyclic graphs can be used to capture the relationships between variables.
In terms of a directed acyclic graph we have

```{r fig-margin1, fig.margin = TRUE, fig.cap = "DAG for speed, colour and type", fig.width=3.5, fig.height=3.5}
# Using Malcolm Barrett's ggdag package
suppressPackageStartupMessages(library(dagitty))
suppressPackageStartupMessages(library(ggdag))

dagified <- dagify(colour ~ type,
                   speed ~ type,
                   exposure = "x",
                   outcome = "speed")
ggdag(dagified, layout = "circle")
```

There are three nodes type, colour and speed. The car type "causes"
the car colour and the car speed so there are arrows from type to
colour and speed.

# Simulating data


To put some probability structure on this let
sports cars and normal cars be equally likely
$$
\begin{array}{rl}
{\mathbb P}( \mbox{sports car}  ) = 0.5 \\
{\mathbb P}( \mbox{normal car}  ) = 0.5 \\
\end{array}
$$
and
$$
\begin{array}{rl}
{\mathbb P}( \mbox{red car} | \mbox{sports car}  ) = 0.8 \\
{\mathbb P}( \mbox{blue car} | \mbox{sports car}  ) = 0.2 \\
{\mathbb P}( \mbox{red car} | \mbox{normal car}  ) = 0.2 \\
{\mathbb P}( \mbox{blue car} | \mbox{normal car}  ) = 0.8 \\
\end{array}
$$
Finally, suppose that the speed of the car is
$$
\mbox{speed} = 100 + 50 \times {\mathbb I}(\mbox{sports car}) + N(0,50^2)
$$

Lets simulate some data that captures this scenario.
```{marginfigure}
__Steph__: I think we might want some side banter.

__John__: What a wonderful idea Steph! I think and equation might be nice.
$$
1 + 1 = 2
$$
```

Let's simulate some data using  [@R-base]

```{r}
prob_sport <- 0.5
prob_red_given_sport  <- 0.8
prob_red_given_normal <- 0.2

mu_sport  <- 150
mu_normal <- 100
sd <- 50

n <- 100
type   <- rep(NA,n)
type_ind <- rbinom(n,1,prob_sport)
type[type_ind==1] <- "sport"
type[type_ind==0] <- "normal"
type <- as.factor(type)

nSport <- sum(type=="sport")
nNormal <- sum(type=="normal")

colour_ind <- rep(NA,n)
colour_ind[type=="sport"]  <- rbinom(nSport,1,prob_red_given_sport)
colour_ind[type=="normal"] <- rbinom(nNormal,1,prob_red_given_normal)

colour <- rep(NA,n)
colour[colour_ind==1] <- "red"
colour[colour_ind==0] <- "blue"
colour <- factor(colour,levels=c("red","blue"))

speed <-  mu_normal + (mu_sport - mu_normal)*(type=="sport") + rnorm(n,sd=50)

suppressPackageStartupMessages(library(tidyverse))

tib <- tibble(type=type,
              colour=colour,
              speed=speed)
```

So if we plot a boxplot for speed against colour we get.

```{r fig-margin2, fig.margin = TRUE, fig.cap = "speed vs colour", fig.width=3.5, fig.height=3.5}
p <- ggplot(tib, aes(x=colour, y=speed, fill=colour)) + 
  geom_boxplot() +
  theme_bw()
p
```

# Boxplots and Two sample t-tests

So if we plot a boxplot speed against type we get.

```{r fig-margin3, fig.margin = TRUE, fig.cap = "speed vs type", fig.width=3.5, fig.height=3.5}
p <- ggplot(tib, aes(x=type, y=speed,fill=type)) + 
  geom_boxplot() +
  theme_bw()
p
```

So if we wanted to see if red cars were significantly faster than
blue cars we could perform a $t$ test.
```{r}
# We use Lukas Burk, Tobias Anton, Daniel Lüdecke, and Gesa Graf's 
#tadaatoolbox package to nicely summarise the results of a two sample t-test.
library(tadaatoolbox)

# Two sample t-test grouped using colour
tadaa_t.test(data = tib, response = speed, group = colour, print = "markdown")
```

We could also do one where we group on type.
```{r}
# Two sample t-test grouped using type.
tadaa_t.test(data = tib, response = speed, group = type, print = "markdown")
```

In both cases we get statistical significance.


# Reproducibility, prediction and inference

Suppose that we went to a number of car yards where data was generated
using the above process and we only observed

* the Speed of the car and the

* the Colour of the car.

Then

* __Reproducibility__: The results would be largely reproducible since
if we went from car yard to car yard red cars would be generally faster
than blue cars. 
Pre-registration does not work.


* __Prediction__: If we were to pick a car that we wanted to be fast
then we would pick a red car and the car would be more likely to be
faster than if we picked a blue car.

* __Inference__: Painting our car red would not make it any faster.


# Post example Banter

__Steph__: Some witty remark


__John__: Ha ha that was so witty Steph.



 
 
```{r bib, include=FALSE}
# create a bib file for the R packages used in this document
knitr::write_bib(c('base', 'rmarkdown'), file = 'BAF.bib')
```
 


<br>

<br>

